---
title: "SplitGenes"
author: "Patrick Monnahan"
date: "9/4/2018"
output: html_document
---

Load packages and Data
```{r setup, include=FALSE}
library(data.table)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(ggridges)
library(parallel)
library(gridExtra)
library(stringr)
library(tibble)
library(magrittr)
library(DESeq2)
library(DEXSeq)
library(wrapr)
library(scales)
library(wesanderson)

#mask these functions from other packages
select <- dplyr::select
filter <- dplyr::filter
mutate <- dplyr::mutate
summarize <- dplyr::summarize

# Load Data
b.splits = read.table("~/Documents/Research/Maize/Split_genes/input/B73_LT_HTseq_splitGene_key.500kb.t0.1.txt", head=T)
b.fsplits = read.table("~/Documents/Research/Maize/Split_genes/input/B73_splitGenes_sim_m4.txt", head=T)
b.counts = read.table("~/Documents/Research/Maize/Split_genes/input/Samples_LT_B73_Ref_HTseq.txt", head=T)
w.splits = read.table("~/Documents/Research/Maize/Split_genes/input/W22_LT_HTseq_splitGene_key.500kb.t0.1.txt",head=T)
w.fsplits = read.table("~/Documents/Research/Maize/Split_genes/input/W22_splitGenes_fake_m4.txt", head=T)
w.counts=read.table("~/Documents/Research/Maize/Split_genes/input/Samples_LT_W22_Ref_HTseq.txt", head=T)
p.splits = read.table("~/Documents/Research/Maize/Split_genes/input/PH207_LT_HTseq_splitGene_key.500kb.t0.1.txt",head=T)
p.fsplits = read.table("~/Documents/Research/Maize/Split_genes/input/PH207_splitGenes_sim_m4.txt",head=T)
p.counts = read.table("~/Documents/Research/Maize/Split_genes/input/Samples_LT_PH207_Ref_HTseq.txt", head=T)

t.splits = read.table("~/Documents/Research/Maize/Split_genes/input/Wsplits_WB_td0_m4.txt", head = T)

#Syntenic genes between B and P
synt = read.table("~/Documents/Research/Maize/MaizeSV/misc/SyntenicGenes_BandP.txt")
```

#Analysis of mean two-fold coverage difference across splitGenes

##Define functions
```{r}
# Convert count data to transcripts per million (tpm)
# Taken from: https://gist.github.com/slowkow/c6ab0348747f86e2748b
counts_to_tpm <- function(counts, featureLength, meanFragmentLength) {
  
  # Ensure valid arguments.
  stopifnot(length(featureLength) == nrow(counts))
  stopifnot(length(meanFragmentLength) == ncol(counts))
  
  # Compute effective lengths of features in each library.
  effLen <- do.call(cbind, lapply(1:ncol(counts), function(i) {
    featureLength - meanFragmentLength[i] + 1
  }))
  
  # Exclude genes with length less than the mean fragment length.
  idx <- apply(effLen, 1, function(x) min(x) > 1)
  counts <- counts[idx,]
  effLen <- effLen[idx,]
  featureLength <- featureLength[idx]
  
  # Process one column at a time.
  tpm <- do.call(cbind, lapply(1:ncol(counts), function(i) {
    rate = log(counts[,i]) - log(effLen[,i])
    denom = log(sum(exp(rate)))
    exp(rate - denom + log(1e6))
  }))
  
  # Copy the row and column names from the original matrix.
  colnames(tpm) <- colnames(counts)
  rownames(tpm) <- rownames(counts)
  return(tpm)
}

# Calculate the average log fold change across a set of potential split genes -- called by calcM2f
getAvgLog = function(v, verbose=F){
  vm = matrix(v, nrow = length(v), ncol = length(v)) # creates a square matrix out of a vector of tpm values
  svm = sweep(vm, 2, v, FUN = "/") # divide rows of matrix by vector (v)...basically get fold change of all relative to all
  lt = abs(log(svm[lower.tri(svm)], 2)) # get upper and lower triangular matrix...i.e. exclude the diagonals of matrix which is just all 1's
  ut = abs(log(svm[upper.tri(svm)], 2)) # abs(log(1/5)) == abs(log(5)).  each fold change gets double counted but this comes out in the average
  m2f = mean(c(lt,ut)) 
  if (verbose==TRUE){
    print(vm)
    print(svm)
    print(lt)
    print(ut)
    print(m2f)
  }
  return(m2f)
}

# Counts number of candidates that exceed the threshold for calling split or merged -- called in plot.m2f
countExceeds2 = function(df, merged_quantile, split_quantile){
  mq = quantile(df[df$source == "simMerged",]$M2f, merged_quantile, na.rm = T)
  sq = quantile(df[df$source == "simSplit",]$M2f, split_quantile, na.rm = T)
  print(sq)
  print(mq)
  counts = df %>%filter(source=="real") %>%mutate(split = ifelse(M2f > sq, 1,0), merged = ifelse(M2f < mq, 1, 0)) 
  
  splits = sum(counts$split, na.rm=T)
  mergeds = sum(counts$merged,na.rm=T)
  # %>% summarize(splits = sum(split,na.rm=T), mergeds = sum(merged,na.rm=T))
  return(c(splits,mergeds))
}

plot.m2f.synt = function(df, merged_quantile, split_quantile){
  n.counts = as.character(countExceeds2(df[df$synt=="nonsynt",], merged_quantile,split_quantile))
  s.counts = as.character(countExceeds2(df[df$synt=="syntenic",], merged_quantile,split_quantile))
  plt = ggplot(df, aes(x=M2f, fill=source, y=synt)) + geom_density_ridges(alpha=0.5,scale=0.9) + geom_vline(xintercept = c(quantile(df[df$source=="simSplit" & df$synt=="syntenic",]$M2f, split_quantile,na.rm=T), quantile(df[df$source=="simMerged" & df$synt=="syntenic",]$M2f, merged_quantile, na.rm=T)), linetype = "longdash") + geom_vline(xintercept = c(quantile(df[df$source=="simSplit" & df$synt=="nonsynt",]$M2f, split_quantile, na.rm=T), quantile(df[df$source=="simMerged" & df$synt=="nonsynt",]$M2f, merged_quantile, na.rm=T)), linetype = "longdash", color="red") + scale_x_log10() + theme_bw() + annotate("text", x = 45.9, y = 1.3, label = paste("Merged =", n.counts[2],"; Split =", n.counts[1])) + annotate("text", x = 1.9, y = 1.4, label = paste("Merged =", s.counts[2],"; Split =", s.counts[1]))
  return(plt)
}

convertFactors = function(df){
  df$tissue = as.factor(df$tissue)
  df$sample = as.factor(df$sample)
  df$rep = as.factor(df$rep)
  df$gene = as.factor(df$gene)
  return(df)
}

parseName = function(df){
  df = df %>%mutate(ref = str_split_fixed(as.character(variable), "_", 3)[,3], sample = str_split_fixed(as.character(variable), "[.]", 3)[,1], tissue = str_split_fixed(as.character(variable), "[.]", 3)[,2], rep = str_split_fixed(as.character(variable), "[.]", 3)[,3]) %>%mutate(rep = str_split_fixed(as.character(rep),"_",3)[,1]) %>% as.data.frame()
  return(df)
}

# Melts combined dataset and labels rows with information in sample_ids
munge = function(df, melt_by){
  d.m = melt(df, id.vars = melt_by)
  d.n = parseName(d.m)
  d.n = convertFactors(d.n)
  return(d.n)
}

#Major function that takes the 3 inputs: exon-based count data from HTseq (counts), file containing candidate split/merge genes(splits), and file contianing the "sim" split/merge gene for use as null distribution
formatData = function(counts, splits, minTPM=0, sampleID = c("B", "P", "W"), minGenes = 1, fmt="de"){
  print("Munging...")
  d = merge(splits, counts, by = c("exon"), all.y=TRUE)
  Names = c("exon","pos.exon","end.exon","gene","pos.gene","end.gene","parent","prog", "source")
  p = munge(d, c(Names, "chrom"))
  #Remove entries without unique exon assignment and retain only the samples that we specify
  p %<>% filter(!exon %in% c("__alignment_not_unique", "__ambiguous", "__no_feature", "__not_aligned", "__too_low_aQual", "__alignment_not_unique")) %>% as.data.frame()
  print("Munging...done")
  p['basepairs'] = abs(p$end.exon - p$pos.exon)
  if (fmt == "m2f"){
    p %<>% filter(sample %in% sampleID) %>% as.data.frame()
    print("Calculating TPM...")
    nse = dcast(p, exon + gene + parent + prog + basepairs + pos.gene + end.gene + pos.exon + end.exon + source ~ variable, value.var = "value")
    nse = nse[nse$basepairs>50,]
    nse=nse[!is.na(nse$basepairs),]
    nse['rowid'] = paste(nse$exon, nse$prog, nse$gene, nse$parent)
    row.names(nse) = nse$rowid
    nse %<>% select(matches(paste(paste(c(Names, "basepairs", "rowid"), collapse = "|"), paste(sampleID, ".", collapse = "|", sep = ""), sep = "|")))
    tpm_e = as.data.frame(counts_to_tpm(nse[, 12:ncol(nse)-1], nse$basepairs, c(rep(50, 20 * length(sampleID)))))
    tpm_e = rownames_to_column(tpm_e, var = "rowid")
    ww = merge(tpm_e, nse[,c(Names, "basepairs", "rowid")], by="rowid", all.x=T)
    print("Done calculating TPM...")
    print("Re-munging...")
    p = munge(ww, c(Names, "basepairs", "rowid"))
    #Average TPM across exons and label parents
    p = p %>% group_by(gene, tissue, sample, rep, source, parent, prog, pos.gene, end.gene) %>% summarize(value = mean(value)) %>% as.data.frame()
    p %<>% mutate(isParent = ifelse(as.character(gene) == as.character(parent), TRUE, FALSE))
    # Stalling here...I still can't figure out why.
    ## Break the mutate section apart...
    # ppp = p %>% group_by(gene, tissue, sample, rep, source, parent, prog, pos.gene, end.gene) %>% mutate(isParent = ifelse(as.character(gene) == as.character(parent), TRUE, FALSE), expressed=ifelse(value!=0,1,0)) %>% summarize(numExp = sum(expressed), numExons = n(), isParent=unique(isParent))
    # p = left_join(pp,ppp,by=c("gene", "tissue", "sample", "rep", "source", "parent", "prog", "pos.gene", "end.gene"))
  }
return(p)
}

calcM2f = function(df, sampleID, minTPM = 0, FUN=mean){
  # Only consider the split/child genes, filter out 0s, unwanted samples
  DF = df %>%filter(isParent == FALSE & value > minTPM & sample == sampleID) %>% group_by(parent, prog, source, sample, rep, tissue) %>% summarize(m2f = getAvgLog(value)) %>% ungroup() %>% group_by(parent, prog, source) %>% summarize(M2f = FUN(m2f,na.rm=T), numTiss=length(unique(as.character(tissue))), BiDiffMean = NaN)
  badPars = DF %>% filter(M2f=="NaN")
  DF2 = df %>%filter(parent %in% badPars$parent & isParent == FALSE & sample == sampleID) %>% mutate(expressed = ifelse(value > 0, 1, 0)) %>% group_by(parent, prog, source, sample, rep, tissue) %>% summarize(diff = mean(abs(diff(expressed)))) %>% ungroup() %>% group_by(parent, prog, source) %>% summarize(M2f = NaN, numTiss=length(unique(as.character(tissue))), BiDiffMean = sum(diff))
  DF %<>% filter(M2f!=0) %>% as.data.frame()
  DF = rbind(DF, as.data.frame(DF2))
  return(DF)
}
 
plot.m2f = function(df, merged_quantile, split_quantile, xmin = 0, xloc = 0.2, yloc = 1.3){
    df %<>% filter(M2f!="-Inf")
    counts = as.character(countExceeds2(df, merged_quantile,split_quantile))
    tot = nrow(df[df$source == "real" & !is.nan(df$M2f),])
    plt = ggplot(df[df$M2f > xmin,], aes(x=M2f, fill=source)) + geom_density(alpha=0.5) + geom_vline(xintercept = quantile(df[df$source=="simSplit",]$M2f, split_quantile, na.rm=T), color = "green", linetype = "longdash") + geom_vline(xintercept = quantile(df[df$source=="simMerged",]$M2f, merged_quantile, na.rm = T), linetype = "longdash", color = "red") + scale_x_log10() + theme_bw() + annotate("text", x = xloc, y = yloc, label = paste("Merge =", counts[2],"\nSplit =", counts[1], "\nTotal =", tot)) + xlab(expression(paste("Mean ", log[2], "(", g[x], "/", g[y], ")")))
    return(plt)
} 

munge6 = function(df, synteny_key){
  df = df %>%mutate(synt = if_else(source == "simMerged", if_else(substr(as.character(parent), 1, nchar(as.character(parent)) - 1) %in% synteny_key$V1,"syntenic","nonsynt"), if_else(parent %in% synteny_key$V1,"syntenic","nonsynt"))) %>% as.data.frame()
  return(df)
}

plot.expression = function(df, parent_id, name=-9, outdir=-9, save=F){
  aa = df %>% filter(parent == parent_id) %.>% ggplot(., aes(x = pos.gene/1000, y = value, shape = sample, color = tissue)) + geom_point(size = 2, position = position_dodge(width = 0.1)) + geom_hline(data = ., aes(yintercept = mean(value)), linetype="dashed") + facet_wrap(~gene, scales="free_x") + xlab("Position (kb)") + ylab("Reads per kb")
  if (save==TRUE){
  ggsave(paste(parent_id,name,".png",sep=""), height=5, width=7, unit="in",plot = last_plot(), device = png(), path = outdir,
  scale = 1,
  dpi = 300, limitsize = TRUE)
  dev.off()
  }
  return(aa)
}

#Adds a column to the data frame labelling whether each call in calcM2f results exceed "significance threshold"
labelExceeds = function(df, merged_quantile, split_quantile, Source = c("real")){
  mq = quantile(df[df$source == "simMerged",]$M2f, merged_quantile, na.rm = T)
  sq = quantile(df[df$source == "simSplit",]$M2f, split_quantile, na.rm = T)
  df %<>%filter(source %in% Source) %>%mutate(Output_call = ifelse(M2f > sq, "SPLIT", ifelse(M2f < mq, "MERGED", "NOCALL"))) %>% as.data.frame() 
  return(df)
}
```


##Calculate m2f
```{r}
#Combine HTseq count data, split gene candidates, and simData candidates
pp = formatData(p.counts, p.splits, p.fsplits, sampleID = "P", fmt = "m2f")
ww = formatData(w.counts, w.splits, w.fsplits, sampleID = "W", fmt = "m2f")
bb = formatData(b.counts, b.splits, b.fsplits, sampleID = "B", fmt = "m2f")


# Calculate the mean log-2-fold expression across splitGene candidates for each set of candidates
P.r = calcM2f(pp, "P", minTPM = 0.01)
B.r = calcM2f(bb, "B", minTPM = 0.01)
W.r = calcM2f(ww, "W", minTPM = 0.01)

W.r['geno']="W"
B.r['geno']="B"
P.r['geno']="P"

ALL = do.call("rbind",list(W.r,P.r,B.r))

#classify candidate split-genes 
upper_quantile = 0.9
lower_quantile = 0.1

ALL %<>% mutate(Call = ifelse(M2f > quantile(ALL[ALL$source=="simSplit",]$M2f, upper_quantile, na.rm=T), "Split", ifelse(M2f < quantile(ALL[ALL$source == "simMerged",]$M2f, lower_quantile, na.rm=T), "Merged", "NoCall")))

ALL %<>% mutate(Par = ifelse(grepl("Zm00001", parent), "B", ifelse(grepl("Zm00008", parent), "P", "W")))
```


# Investigate effects of using different functions to summarize instead of just Mean
```{r}
W.r.max = calcM2f(ww, "W", minTPM = 0.01,FUN=max)
W.r.med = calcM2f(ww, "W", minTPM = 0.01,FUN=median)
W.r.var = calcM2f(ww, "W", minTPM = 0.01,FUN=var)
plot.m2f(W.r, 0.1, 0.9, xmin=0.1)
plot.m2f(W.r.max, 0.1, 0.9, xmin=0.1)
plot.m2f(W.r.med, 0.1, 0.9, xmin=0.1)
plot.m2f(W.r.var, 0.1, 0.9, xmin=0.1)

WW.r = labelExceeds(W.r, 0.1, 0.9)
WW.r.max = labelExceeds(W.r.max , 0.1, 0.9)
WW.r.med = labelExceeds(W.r.med, 0.1, 0.9)
WW.r.var = labelExceeds(W.r.var, 0.1, 0.9)

WW = merge(WW.r,WW.r.max,by=c("parent","prog","source"),suffixes=c(".mean",".max"))
WW = merge(WW,WW.r.med,by=c("parent","prog","source"))
WW = merge(WW,WW.r.var,by=c("parent","prog","source"),suffixes=c(".med",".var"))

WW %>% group_by(Output_call.mean, Output_call.max, Output_call.med, Output_call.var) %>% summarize(n=n()) %>% as.data.frame()
```


# Candidate summary
```{r}
#What about filter for genes that are reciprical one-one across all 3 comparisons?
oo = read.table("~/Documents/Research/Maize/Split_genes/data/One_to_ones.txt")
oo['gene'] = oo$V1
oo %<>% group_by(V1,V2) %>% mutate(gene1 = min(as.character(V1),as.character(V2)), gene2 = max(as.character(V1),as.character(V2))) # Sort the one-to-one gene ids so that we can find reciprocals as pairs of matching gene ids

OO3 = oo %>% group_by(gene1) %>% summarize(n=n()) %>% filter(n==4) %>% mutate(gene = gene1) %>% select(gene)
OO4 = oo %>% group_by(gene2) %>% summarize(n=n()) %>% filter(n==4) %>% mutate(gene = gene2) %>% select(gene)
OO5 = rbind(OO3, OO4)

cMerge = ALL %>% filter(source=="real" & Call == "Merged")
cSplits = ALL %>% filter(source=="real" & Call == "Split")

bb["ref"]="B73"
ww['ref']="W22"
pp['ref']="PH207"

all = do.call("rbind", list(bb,ww,pp))

all %<>% mutate(length = end.gene - pos.gene, propExp = numExp / numExons)

all %<>% mutate(ISPAR = ifelse(gene %in% bb[bb$source=="real",]$parent, TRUE, ifelse(gene %in% pp[pp$source=="real",]$parent, TRUE, ifelse(gene %in% ww[ww$source=="real",]$parent,TRUE,FALSE))))

all %<>% mutate(gene.ref = case_when(substr(gene, 7, 7) == "1" ~ "B", substr(gene, 7, 7) == "8" ~ "P", substr(gene, 7, 7) == "4" ~ "W"))

all %<>% mutate(called_split = ifelse(parent %in% cSplits$parent, "split", ifelse(parent %in% cMerge$parent, "merge", "nocall")))
all$called_split = as.factor(all$called_split)

all %<>% mutate(Called_split = paste(called_split,ISPAR))

all$called_split=factor(all$called_split,levels(all$called_split)[c(3, 2, 1)])

#Isolate one to one genes
all.11 = all %>% filter(called_split =="nocall" & gene %in% OO6$gene) %>% mutate(called_split = "One-to-one")

#Isolate just the classified split and merged info
all2 = all %>% filter(source %in% c("real","unchanged") & called_split %in% c("split","merge"))
all2 = rbind(all2, all.11)

all2$called_split=factor(all2$called_split)
all2$called_split=factor(all2$called_split,levels(all2$called_split)[c(1, 3, 2)])

all2 %<>% mutate(State = paste(called_split,ISPAR, sep=""))

all2 %<>% mutate(State = replace(State, grepl("One-to-one", State), "One-to-one"))

all2$State=factor(all2$State)
all2$State=factor(all2$State,levels(all2$State)[c(3, 4, 1, 2, 5)])

#How many genes are we encompassed by our candidates??
all %>% filter(source=="real") %>% group_by(ref) %>% distinct(gene) %>% summarize(n=n())

# Gene length plot
outs = all2 %>% group_by(gene.ref, State) %>% mutate(outlier = (log(abs(length)) > median(log(abs(length))) + IQR(log(abs(length))) * 1.5) | (log(abs(length)) < median(log(abs(length))) - IQR(log(abs(length))) * 1.5)) %>% distinct(gene.ref,parent, prog, State, outlier, length) %>% filter(outlier==T)

#Outlier jitter is still not working
all2 %>% select(source,prog,ISPAR,length, State, gene.ref) %>% distinct() %>% ggplot(.,aes(x=gene.ref,fill=State,y=abs(length)/1000)) + geom_boxplot(outlier.shape=NA) + geom_point(data = outs, aes(x=gene.ref,fill=State,y=abs(length)/1000), alpha = 0.2, position=position_jitterdodge(dodge.width=0.8)) + scale_y_log10(label=comma) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + xlab("Annotation") + ylab("Gene length (kb)") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

#no outliers
all2 %>% select(source,prog,ISPAR,length, State, gene.ref) %>% distinct() %>% ggplot(.,aes(x=gene.ref,fill=State,y=abs(length)/1000)) + geom_boxplot(outlier.shape=NA) + scale_y_log10(label=comma, limits=c(0.1,100)) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + xlab("Annotation") + ylab("Gene length (kb)") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

#Exon number
all2 %>% select(gene.ref,source,prog,ISPAR,sample,numExons, State) %>% distinct() %>% ggplot(.,aes(x=gene.ref,fill=State,y=numExons)) + geom_boxplot(outlier.shape=NA) + scale_y_log10(label=comma) + scale_fill_manual(name="",labels=c("1-1", "SC", "SNC", "MC","MNC"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + xlab("Annotation") + ylab("Exon Number") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

#Exon Density
all2 %>% select(gene.ref,source,prog,ISPAR,sample,numExons, length, State) %>% distinct() %>% ggplot(.,aes(x=gene.ref,fill=State,y=numExons/ abs(length))) + geom_boxplot(outlier.shape=NA) + scale_y_log10() + scale_fill_manual(name="",labels=c("1-1", "SC", "SNC", "MC","MNC"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + xlab("Annotation") + ylab("Exon Density") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

#Proportion of exons that are expressed
all2 %>% select(source,prog,ISPAR,sample,propExp, State) %>% distinct() %>% ggplot(.,aes(x=sample,fill=State,y=propExp)) + geom_boxplot(outlier.shape=NA) + scale_y_log10(label=comma) + scale_fill_manual(name="Annot./Called",labels=c("1-1", "SC", "SNC", "MC","MNC"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + xlab("Annotation") + ylab("Proportion Exons Expressed") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

# Expression distributions
all2 %>% filter(value!=0) %>% select(gene.ref,source,length,numExp,numExons,propExp,State,value) %.>% ggplot(., aes(y=gene.ref,x=value,fill=State)) + geom_density_ridges(alpha=0.5,scale=0.99) + scale_x_log10(labels=comma) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + xlab("Transcripts per Million") + ylab("Annotation") + theme_bw()+ theme(axis.text.y=element_text(size=16), axis.title.y = element_text(size=18), axis.text.x=element_text(size=16), axis.title.x=element_text(size=18), legend.text = element_text(size=16), legend.title = element_text(size=18))

all2 %>% filter(value!=0 & State %in% c("splitTRUE","One-to-one","mergeFALSE")) %>% select(gene.ref,source,length,numExp,numExons,propExp,State,value) %.>% ggplot(., aes(y=gene.ref,x=value,fill=State)) + geom_density_ridges(alpha=0.5,scale=0.99) + scale_x_log10(labels=comma) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + xlab("Transcripts per Million") + ylab("Annotation") + theme_bw()+ theme(axis.text.y=element_text(size=16), axis.title.y = element_text(size=18), axis.text.x=element_text(size=16), axis.title.x=element_text(size=18), legend.text = element_text(size=16), legend.title = element_text(size=18))

#Number of tissues in which gene is expressed
#Not convinced that the outlier plotting is going to work.  Just too messy. Can just report number of instances of < 3 tissues for each category.
outs = all2 %>% filter(value>0.1) %>% group_by(gene.ref,gene, parent, prog, called_split, sample, source, ISPAR, State) %>% summarize(numTiss = n() / 2) %>% ungroup() %>% group_by(gene.ref, State) %>% mutate(outlier = numTiss < quantile(numTiss, 0.25) - IQR(numTiss) * 1.5) %>% distinct(gene.ref,parent, prog, State, outlier, numTiss)

all2 %>% filter(value>0.1) %>% group_by(gene.ref,gene, parent, prog, called_split, sample, source, ISPAR, State) %>% summarize(numTiss = n() / 2) %>% ggplot(.,aes(x=gene.ref,fill=State,y=numTiss)) + geom_boxplot(outlier.shape=NA) + geom_point(data = outs, aes(x=gene.ref,fill=State,y=numTiss, alpha=outlier), position=position_jitterdodge(dodge.width=0.63)) + scale_alpha_discrete(guide = F, range=c(0,0.2)) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + xlab("Annotation") + ylab("Tissues w/ Expression") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

all2 %>% filter(value>0.1) %>% group_by(gene.ref,gene, parent, prog, called_split, sample, source, ISPAR, State) %>% summarize(numTiss = n() / 2) %>% ggplot(.,aes(x=gene.ref,fill=State,y=numTiss)) + geom_boxplot(outlier.shape=NA) + geom_point(data = outs, aes(x=gene.ref,fill=State,y=numTiss, alpha=outlier,color=State), position=position_jitterdodge(dodge.width=0.63)) + scale_alpha_discrete(guide = F, range=c(0,0.4)) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + scale_color_manual(guide=F,name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + xlab("Annotation") + ylab("Tissues w/ Expression") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

all2 %>% filter(value>0.1) %>% group_by(gene.ref,gene, parent, prog, called_split, sample, source, ISPAR, State) %>% summarize(numTiss = n() / 2) %>% ggplot(.,aes(x=gene.ref,fill=State,y=numTiss)) + geom_boxplot(outlier.shape=NA) + geom_point(data = outs, aes(x=gene.ref,fill=State,y=numTiss, alpha=outlier,color=State), position=position_jitterdodge(dodge.width=0.63)) + scale_alpha_discrete(guide = F, range=c(0,0.4)) + scale_fill_manual(name="",labels=c("1-1", "SC", "SNC", "MC","MNC"),values=wes_palette("Royal2")) + scale_color_manual(guide=F,name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + xlab("Annotation") + ylab("Tissues w/ Expression") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

all2 %>% filter(value!=0) %>% group_by(gene.ref,gene, parent, prog, called_split, sample, source, ISPAR, State) %>% summarize(numTiss = n() / 2) %>% ggplot(.,aes(x=gene.ref,fill=State,y=numTiss)) + geom_boxplot(outlier.shape=NA) + scale_y_log10(label=comma) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + xlab("Annotation") + ylab("Number of Tissues w/ Expression") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

#Prop genes with expression in < 5 tissues
all2 %>% filter(value > 0.01) %>% group_by(gene.ref,gene, parent, prog, called_split, sample, source, ISPAR, State) %>% summarize(numTiss = n() / 2) %>% group_by(gene.ref, State) %>% mutate(outlier = numTiss < 3) %>% distinct(gene.ref,parent, prog, State, outlier, numTiss) %>% summarize(outs = sum(outlier), prop = sum(outlier) / n()) %>% ggplot(., aes(x = gene.ref, fill = State, y = prop)) + geom_bar(stat="identity", position = "dodge") + scale_fill_manual(name="",labels=c("1-1", "SC", "SNC", "MC","MNC"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + xlab("Annotation") + ylab("Prop. genes expressed in < 5 tissues") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

#Prop tissues
all2 %>% filter(value!=0) %>% group_by(gene.ref,gene, parent, prog, called_split, sample, source, ISPAR, State) %>% summarize(numTiss = n() / 2) %>% ggplot(.,aes(x=gene.ref,fill=State,y=numTiss / 10)) + geom_boxplot(outlier.shape=NA) + scale_y_continuous(limits=c(0,0.8)) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + xlab("Annotation") + ylab("Number of Tissues w/ Expression") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

all2 %>% filter(value!=0) %>% group_by(gene.ref,gene, parent, prog, called_split, sample, source, ISPAR, State) %>% summarize(numTiss = n() / 2) %>% ggplot(.,aes(y=gene.ref,fill=State,x=numTiss/10)) + geom_density_ridges(alpha = 0.7, scale=0.99) + scale_y_log10(label=comma) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + xlab("Annotation") + ylab("Prop. Tissues w/ Expression") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))


#Distance between genes
p.dist = p.fsplits %>% filter(source=="unchanged") %>% distinct(chrom,pos.gene,end.gene,gene,source,parent) %>% mutate(sort="group2")
p.dist2 = p.splits %>% distinct(chrom,pos.gene,end.gene,gene,parent) %>% mutate(source="real",sort="group1") %>% rbind(.,p.dist)
P.dist = p.fsplits %>% filter(gene %in% b.splits$parent | gene %in% w.splits$parent) %>% distinct(chrom,pos.gene,end.gene,gene,parent) %>% mutate(source="merged",sort="group2") %>% rbind(.,p.dist2)

b.dist = b.fsplits %>% filter(source=="unchanged") %>% distinct(chrom,pos.gene,end.gene,gene,source,parent) %>% mutate(sort="group2")
b.dist2 = b.splits %>% distinct(chrom,pos.gene,end.gene,gene,parent) %>% mutate(source="real",sort="group1") %>% rbind(.,b.dist)
B.dist = b.fsplits %>% filter(gene %in% p.splits$parent | gene %in% w.splits$parent) %>% distinct(chrom,pos.gene,end.gene,gene,parent) %>% mutate(source="merged",sort="group2") %>% rbind(.,b.dist2)

w.dist = w.fsplits %>% filter(source=="unchanged") %>% distinct(chrom,pos.gene,end.gene,gene,source,parent) %>% mutate(sort="group2")
w.dist2 = w.splits %>% distinct(chrom,pos.gene,end.gene,gene,parent) %>% mutate(source="real",sort="group1") %>% rbind(.,w.dist)
W.dist = w.fsplits %>% filter(gene %in% p.splits$parent | gene %in% b.splits$parent) %>% distinct(chrom,pos.gene,end.gene,gene,parent) %>% mutate(source="merged",sort="group2") %>% rbind(.,w.dist2)

W.dist %<>% arrange(sort,chrom,pos.gene) %>% mutate(diff = pmin(abs(pos.gene - lag(end.gene)), abs(end.gene - lead(pos.gene))))
P.dist %<>% arrange(sort,chrom,pos.gene) %>% mutate(diff = pmin(abs(pos.gene - lag(end.gene)), abs(end.gene - lead(pos.gene))))
B.dist %<>% arrange(sort,chrom,pos.gene) %>% mutate(diff = pmin(abs(pos.gene - lag(end.gene)), abs(end.gene - lead(pos.gene))))

dist = do.call("rbind", list(W.dist, P.dist, B.dist))

dist %<>% mutate(gene.ref = case_when(substr(gene, 7, 7) == "1" ~ "B", substr(gene, 7, 7) == "8" ~ "P", substr(gene, 7, 7) == "4" ~ "W"))
dist$source=as.factor(dist$source)
dist$source=factor(dist$source,levels(dist$source)[c(2,3,1)])

dist %<>% mutate(called_split = ifelse(gene %in% cSplits$gene, "split", ifelse(parent %in% cMerge$parent, "merge", "nocall")))
dist$called_split = as.factor(dist$called_split)

dist %<>% mutate(ISPAR = ifelse(gene %in% bb[bb$source=="real",]$parent, TRUE, ifelse(gene %in% pp[pp$source=="real",]$parent, TRUE, ifelse(gene %in% ww[ww$source=="real",]$parent,TRUE,FALSE))))

#Isolate one-to-ones
dist.11 = dist %>% filter(gene %in% oo$V1 | gene %in% oo$V2) %>% mutate(called_split = "One-to-one")

#Isolate candidates
dist2 = dist %>% filter(source %in% c("real","unchanged") & called_split %in% c("split","merge"))

dist2 = rbind(dist2, dist.11)

dist2$called_split=factor(dist2$called_split)
dist2$called_split=factor(dist2$called_split,levels(dist2$called_split)[c(3,2,1)])

dist2 %<>% mutate(State = paste(called_split,ISPAR, sep=""))

dist2 %<>% mutate(State = replace(State, grepl("One-to-one", State), "One-to-one"))

dist2$State=factor(dist2$State)
dist2$State=factor(dist2$State,levels(dist2$State)[c(3, 4, 1, 2, 5)])

dist2 %>% select(gene.ref,source,ISPAR,diff, State) %>% distinct() %>% ggplot(., aes(y = diff/1000, x = gene.ref, fill = State)) + geom_boxplot(outlier.shape = NA) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P", "W"), labels = c("B73", "PH207", "W22")) + scale_y_log10(label=comma) + xlab("Annotation") + ylab("Distance to nearest gene (kb)") + theme_bw()+ theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))


# Number of genes with no expression
#This version does not throw out 0's
formatData = function(counts,splits,fsplits, minTPM=0, sampleID = c("B", "P", "W"), minGenes = 1, fmt="de"){
  splits['source']="real"
  print("Munging...")
  splits = rbind(splits, fsplits)
  d = merge(splits, counts, by = c("exon"), all.y=TRUE)
  names = c("exon","chrom","pos.exon","end.exon","gene","pos.gene","end.gene","parent","prog", "source")
  p = munge(d, names)
  #Remove entries without unique exon assignment and retain only the samples that we specify
  p %<>% filter(!exon %in% c("__alignment_not_unique", "__ambiguous", "__no_feature", "__not_aligned", "__too_low_aQual", "__alignment_not_unique")) %>% as.data.frame()
  print("Munging...done")
  p['basepairs'] = abs(p$end.exon - p$pos.exon)
  if (fmt == "m2f"){
    p %<>% filter(sample %in% sampleID) %>% as.data.frame()
    print("Calculating TPM...")
    nse = dcast(p, exon + gene + parent + prog + basepairs + pos.gene + end.gene + pos.exon + end.exon + source ~ variable, value.var = "value")
    nse = nse[nse$basepairs>50,]
    nse=nse[!is.na(nse$basepairs),]
    nse['rowid'] = paste(nse$exon, nse$prog, nse$gene, nse$parent)
    row.names(nse) = nse$rowid
    nse %<>% select(matches(paste("exon", "gene", "parent", "prog", "basepairs", 
      "pos.gene", "end.gene", "pos.exon", "end.exon", "source", "rowid", paste(sampleID, ".", collapse = "|", sep = ""), sep = "|")))
    tpm_e = as.data.frame(counts_to_tpm(nse[, 13:ncol(nse)-1], nse$basepairs, c(rep(50, 20 * length(sampleID)))))
    tpm_e = rownames_to_column(tpm_e, var = "rowid")
    ww = merge(tpm_e, nse[,c("exon","gene","parent","prog","basepairs","pos.gene","end.gene","pos.exon","end.exon","source","rowid")], by="rowid", all.x=T)
    print("Done calculating TPM...")
    print("Re-munging...")
    p = munge(ww, c("exon","gene","parent","prog", "source","pos.exon","end.exon","rowid","basepairs","pos.gene","end.gene"))
    #Average TPM across exons and label parents
    pp = p %>% group_by(gene, tissue, sample, rep, source, parent, prog, pos.gene, end.gene) %>% summarize(value = mean(value)) %>% mutate(isParent = ifelse(as.character(gene) == as.character(parent), TRUE, FALSE)) %>% as.data.frame()
    # pp = p %>% group_by(gene, tissue, sample, rep, source, parent, prog, pos.gene, end.gene) %>% filter(value!=0) %>% summarize(value = mean(value)) %>% mutate(isParent = ifelse(as.character(gene) == as.character(parent), TRUE, FALSE)) %>% as.data.frame()
    ppp = p %>% group_by(gene, tissue, sample, rep, source, parent, prog, pos.gene, end.gene) %>% mutate(expressed=ifelse(value!=0,1,0)) %>% summarize(numExp = sum(expressed), numExons = n())
    pp %<>% left_join(.,ppp,by=c("gene", "tissue", "sample", "rep", "source", "parent", "prog", "pos.gene", "end.gene"))
  }
return(pp)
}

pp2 = formatData(p.counts, p.splits, p.fsplits, sampleID = "P", fmt = "m2f")
ww2= formatData(w.counts, w.splits, w.fsplits, sampleID = "W", fmt = "m2f")
bb2 = formatData(b.counts, b.splits, b.fsplits, sampleID = "B", fmt = "m2f")

bb2["ref"]="B73"
ww2['ref']="W22"
pp2['ref']="PH207"

all2 = rbind(bb2,ww2)
all2 = rbind(all2,pp2)

all2["length"] = all2$end.gene - all2$pos.gene

all2 %<>% mutate(ISPAR = ifelse(gene %in% bb2[bb2$source=="real",]$parent, TRUE, ifelse(gene %in% pp2[pp2$source=="real",]$parent, TRUE, ifelse(gene %in% ww2[ww2$source=="real",]$parent,TRUE,FALSE))))

# Number of genes with no expression
ww2 %>% group_by(source,prog) %>% summarize(M2f = mean(value)) %>% ungroup() %>% group_by(source) %>% mutate(noExp = ifelse(M2f==0,1,0)) %>% summarize(noExp = sum(noExp), n=n(), prop = sum(noExp)/n())
bb2 %>% group_by(source,prog) %>% summarize(M2f = mean(value)) %>% ungroup() %>% group_by(source) %>% mutate(noExp = ifelse(M2f==0,1,0)) %>% summarize(noExp = sum(noExp), n=n(), prop = sum(noExp)/n())
pp2 %>% group_by(source,prog) %>% summarize(M2f = mean(value)) %>% ungroup() %>% group_by(source) %>% mutate(noExp = ifelse(M2f==0,1,0)) %>% summarize(noExp = sum(noExp), n=n(), prop = sum(noExp)/n())

#M2f does not correlate strongly with gene length which should be decent evidence that 3' sequencing bias is not driving results
ALL['gene']=ALL$parent
tt = merge(ALL,all[c("gene","length","numExons")], by="gene")
ggplot(tt,aes(x=length,y=M2f)) + geom_point()
```

#M2f distributions
```{r}
ALL %>% filter(source=="real" & !is.na(Call)) %>% group_by(geno, Call) %>% summarize(n=n())
ALL %>% filter(source=="real" & !is.na(Call)) %>% group_by(geno) %>% summarize(n=n())

ALL %>% filter(source=="real") %>% ggplot(.,aes(fill=geno,x=M2f))+geom_density(alpha=0.5)+scale_x_log10(limits=c(0.1,10)) + theme_bw() + scale_fill_manual(name="Genotype",values=wes_palette("GrandBudapest1"))+ geom_vline(xintercept = quantile(ALL[ALL$source=="simSplit",]$M2f,0.9,na.rm=T),linetype="dashed") + geom_vline(xintercept = quantile(ALL[ALL$source=="simMerged",]$M2f,0.1,na.rm=T),linetype="dashed") + ylim(0,1.4) + ylab("") + theme(axis.text.y=element_blank(),axis.text.x=element_text(size=14),axis.title.x=element_text(size=16),legend.text = element_text(size=14))
ALL %>% filter(source %in% c("real")) %>% ggplot(.,aes(fill=source,x=M2f))+geom_density(alpha=0.5)+scale_x_log10(limits=c(0.1,10)) + theme_bw() + scale_fill_manual(name="",labels=c("All Candidates","Sim. Splits", "Sim. Merged"),values=wes_palette("Darjeeling1")) + ylim(0,1.4) 
ALL %>% filter(source %in% c("real","simSplit")) %>% ggplot(.,aes(fill=source,x=M2f))+geom_density(alpha=0.5)+scale_x_log10(limits=c(0.1,10)) + theme_bw() + scale_fill_manual(name="",labels=c("All Candidates","Sim. Splits", "Sim. Merged"),values=wes_palette("Darjeeling1")) + ylim(0,1.4)
ALL %>% filter(source %in% c("real","simSplit")) %>% ggplot(.,aes(fill=source,x=M2f))+geom_density(alpha=0.5)+scale_x_log10(limits=c(0.1,10)) + theme_bw() + scale_fill_manual(name="",labels=c("All Candidates","Sim. Splits", "Sim. Merged"),values=wes_palette("Darjeeling1"))+ geom_vline(xintercept = quantile(ALL[ALL$source=="simSplit",]$M2f,0.9,na.rm=T),linetype="dashed") + ylim(0,1.4)
ALL %>% filter(source %in% c("real","simSplit","simMerged")) %>% ggplot(.,aes(fill=source,x=M2f))+geom_density(alpha=0.5)+scale_x_log10(limits=c(0.1,10)) + theme_bw() + scale_fill_manual(name="",labels=c("All Candidates","Sim. Splits", "Sim. Merged"),values=wes_palette("Darjeeling1"))+ geom_vline(xintercept = quantile(ALL[ALL$source=="simSplit",]$M2f,0.9,na.rm=T),linetype="dashed") + ylim(0,1.4)
ALL %>% filter(source %in% c("real","simSplit","simMerged")) %>% ggplot(.,aes(fill=source,x=M2f))+geom_density(alpha=0.5)+scale_x_log10(limits=c(0.1,10)) + theme_bw() + scale_fill_manual(name="",labels=c("All Candidates","Sim. Splits", "Sim. Merged"),values=wes_palette("Darjeeling1")) + geom_vline(xintercept = quantile(ALL[ALL$source=="simSplit",]$M2f,0.9,na.rm=T),linetype="dashed") + geom_vline(xintercept = quantile(ALL[ALL$source=="simMerged",]$M2f,0.1,na.rm=T),linetype="dashed") + ylim(0,1.4)
ALL %>% filter(source %in% c("real","simMerged")) %>% ggplot(.,aes(fill=source,x=M2f))+geom_density(alpha=0.5)+scale_x_log10(limits=c(0.1,10)) + theme_bw() + scale_fill_manual(name="",labels=c("All Candidates", "Sim. Merged"),values=wes_palette("Darjeeling1")) + ylim(0,1.4)
ALL %>% filter(source %in% c("real","simMerged")) %>% ggplot(.,aes(fill=source,x=M2f))+geom_density(alpha=0.5)+scale_x_log10(limits=c(0.1,10)) + theme_bw() + scale_fill_manual(name="",labels=c("All Candidates", "Sim. Merged"),values=wes_palette("Darjeeling1")) + geom_vline(xintercept = quantile(ALL[ALL$source=="simMerged",]$M2f,0.1,na.rm=T),linetype="dashed") + ylim(0,1.4)
ALL %>% filter(source %in% c("simSplit")) %>% ggplot(.,aes(fill=source,x=M2f))+geom_density(alpha=0.5)+scale_x_log10(limits=c(0.1,10)) + theme_bw() + scale_fill_manual(name="",labels=c("Sim. Splits", "Sim. Merged"),values=wes_palette("Moonrise3")) + ylim(0,1.4) + ylab("") + theme(axis.text.y=element_blank(),axis.text.x=element_text(size=14),axis.title.x=element_text(size=16),legend.text = element_text(size=14))
ALL %>% filter(source %in% c("simSplit")) %>% ggplot(.,aes(fill=source,x=M2f))+geom_density(alpha=0.5)+scale_x_log10(limits=c(0.1,10)) + theme_bw() + scale_fill_manual(name="",labels=c("Sim. Splits", "Sim. Merged"),values=wes_palette("Moonrise3"))+ geom_vline(xintercept = quantile(ALL[ALL$source=="simSplit",]$M2f,0.9,na.rm=T),linetype="dashed") + ylim(0,1.4) + ylab("") + theme(axis.text.y=element_blank(),axis.text.x=element_text(size=14),axis.title.x=element_text(size=16),legend.text = element_text(size=14))
ALL %>% filter(source %in% c("simSplit","simMerged")) %>% ggplot(.,aes(fill=source,x=M2f))+geom_density(alpha=0.5)+scale_x_log10(limits=c(0.1,10)) + theme_bw() + scale_fill_manual(name="",labels=c("Sim. Splits", "Sim. Merged"),values=wes_palette("Moonrise3"))+ geom_vline(xintercept = quantile(ALL[ALL$source=="simSplit",]$M2f,0.9,na.rm=T),linetype="dashed") + ylim(0,1.4) + ylab("") + theme(axis.text.y=element_blank(),axis.text.x=element_text(size=14),axis.title.x=element_text(size=16),legend.text = element_text(size=14))
ALL %>% filter(source %in% c("simSplit","simMerged")) %>% ggplot(.,aes(fill=source,x=M2f))+geom_density(alpha=0.5)+scale_x_log10(limits=c(0.1,10)) + theme_bw() + scale_fill_manual(name="",labels=c("Sim. Splits", "Sim. Merged"),values=wes_palette("Moonrise3")) + geom_vline(xintercept = quantile(ALL[ALL$source=="simSplit",]$M2f,0.9,na.rm=T),linetype="dashed") + geom_vline(xintercept = quantile(ALL[ALL$source=="simMerged",]$M2f,0.1,na.rm=T),linetype="dashed") + ylim(0,1.4) + ylab("") + theme(axis.text.y=element_blank(),axis.text.x=element_text(size=14),axis.title.x=element_text(size=16),legend.text = element_text(size=14))
ALL %<>% mutate(Call = ifelse(M2f > quantile(ALL[ALL$source=="simSplit",]$M2f,0.9,na.rm=T), "Split", ifelse(M2f < quantile(ALL[ALL$source=="simMerged",]$M2f,0.1,na.rm=T), "Merged", "NoCall")))

ALL %>% filter(source %in% c("simMerged")) %>% ggplot(.,aes(x = M2f, fill = "#F4B5BD")) + geom_density(alpha = 0.5) + scale_x_log10(limits = c(0.1,10)) + theme_bw() + geom_vline(xintercept = quantile(ALL[ALL$source=="simMerged",]$M2f,0.1,na.rm=T),linetype="dashed") + ylim(0,1.4) + ylab("") + theme(axis.text.y=element_blank(),axis.text.x=element_text(size=14),axis.title.x=element_text(size=16), legend.position="NA",legend.text = element_blank())

ALL$Call=as.factor(ALL$Call)
ALL$Call=factor(ALL$Call,levels(ALL$Call)[c(2,1,3)])
ALL %<>% mutate(Par = ifelse(grepl("Zm00001", parent),"B",ifelse(grepl("Zm00008",parent),"P","W")))
ALL %>% filter(source=="real" & !is.na(Call)) %>% group_by(Par,geno,Call) %>% summarize(n=n()) %>% ggplot(.,aes(x=Call,fill=Call,y=n))+geom_bar(stat="identity") + facet_grid(rows=vars(geno),cols = vars(Par)) + xlab("") + theme_bw()+ theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) + theme(axis.text.x=element_blank(), axis.title.y = element_text(size=18), axis.text.y=element_text(size=16), legend.title=element_text(size=18), legend.text = element_text(size=16), axis.ticks.x=element_blank(), strip.text = element_text(size=16)) + scale_fill_manual(values = c("#F8766D", "#00BA38", "#619CFF"))
```


#IsoSeq analysis
```{r}
#Require reciprocal overlap of 0.75 for both isoseq reads and genes.  prevents overly long reads that span a gene but don't actually align to it at all.  
iso3 = read.table("~/Documents/Research/Maize/B73_isoseq/IsoSeq.sa.mq20.Ff75.counts")
iso3['parent'] = iso3$V4
iso3['gene'] = iso3$V4
ALL %<>% left_join(.,iso3,by="parent")

ALL %<>% mutate(ISPAR = ifelse(gene %in% bb[bb$source=="real",]$parent, TRUE, ifelse(gene %in% pp[pp$source=="real",]$parent, TRUE, ifelse(gene %in% ww[ww$source=="real",]$parent,TRUE,FALSE))))


ball = all2
b.pars = ball %>% filter(grepl("Zm00001", parent))
b.prog = ball %>% filter(grepl("Zm00001", prog))
b.pars %<>% left_join(.,iso3[,-7],by="parent")
b.prog %<>% left_join(.,iso3[,-6],by="gene")
ball = rbind(b.pars,b.prog)

ball %>% filter(gene.ref == "B") %>% select(gene.ref, gene,parent, prog,source,V5, State) %>% distinct() %>% ggplot(.,aes(x=gene.ref,fill=State,y=V5)) + geom_boxplot(outlier.shape=NA) + scale_y_log10(label=comma) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + xlab("Annotation") + scale_x_discrete(breaks = "B", labels = "B73") + ylab("Number of IsoSeq Reads") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

ball %>% filter(gene.ref == "B") %>% mutate(zero = ifelse(V5==0,1,0)) %>% select(gene,V5, State, zero) %>% distinct() %>% group_by(State) %>% summarize(z = sum(zero,na.rm=T), prop = sum(zero,na.rm=T) / n()) %>% ggplot(.,aes(x = "B",fill = State, y = prop)) + geom_bar(stat="identity", position = "dodge") + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = "B", labels = "B73") + xlab("Annotation") + ylab("Prop. Genes w/o IsoSeq Support") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))


```



Look for effect of minimum exon number flag in simulating splitMerges.  Also consider effect of multiple isoforms
```{r}
p.fsplits20 = read.table("~/Documents/Research/Maize/Split_genes/input/PH207_splitGenes_sim_m20_all.txt",head=T)
b.fsplits20 = read.table("~/Documents/Research/Maize/Split_genes/input/B73_splitGenes_sim_m20_all.txt", head=T)
w.fsplits20 = read.table("~/Documents/Research/Maize/Split_genes/input/W22_splitGenes_sim_m20_all.txt", head=T)

p.fsplits2 = read.table("~/Documents/Research/Maize/Split_genes/input/PH207_splitGenes_sim_m2_all.txt",head=T)
b.fsplits2 = read.table("~/Documents/Research/Maize/Split_genes/input/B73_splitGenes_sim_m2_all.txt", head=T)
w.fsplits2 = read.table("~/Documents/Research/Maize/Split_genes/input/W22_splitGenes_sim_m2_all.txt", head=T)

mi = read.table("~/Documents/Research/Maize/references/multiIsoform_B73.txt")

pp20 = formatData(p.counts, p.splits, p.fsplits20, sampleID = "P", fmt = "m2f")
ww20 = formatData(w.counts, w.splits, w.fsplits20, sampleID = "W", fmt = "m2f")
bb20 = formatData(b.counts, b.splits, b.fsplits20, sampleID = "B", fmt = "m2f")

P.r20 = calcM2f(pp20, "P", minTPM = 0.01)
B.r20 = calcM2f(bb20, "B", minTPM = 0.01)
W.r20 = calcM2f(ww20, "W", minTPM = 0.01)

pp2 = formatData(p.counts, p.splits, p.fsplits2, sampleID = "P", fmt = "m2f")
ww2 = formatData(w.counts, w.splits, w.fsplits2, sampleID = "W", fmt = "m2f")
bb2 = formatData(b.counts, b.splits, b.fsplits2, sampleID = "B", fmt = "m2f")

P.r2 = calcM2f(pp2, "P", minTPM = 0.01)
B.r2 = calcM2f(bb2, "B", minTPM = 0.01)
W.r2 = calcM2f(ww2, "W", minTPM = 0.01)

B.r['min.exons'] = 4
P.r['min.exons'] = 4
W.r['min.exons'] = 4
B.r['ref'] = "B"
P.r['ref'] = "P"
W.r['ref'] = "W"
B.r2['ref'] = "B"
P.r2['ref'] = "P"
W.r2['ref'] = "W"
B.r2['min.exons'] = 2
P.r2['min.exons'] = 2
W.r2['min.exons'] = 2
W.r20['ref'] = "W"
B.r20['ref'] = "B"
P.r20['ref'] = "P"
W.r20['min.exons'] = 20
B.r20['min.exons'] = 20
P.r20['min.exons'] = 20
# W.r8['ref'] = "W"
# B.r8['ref'] = "B"
# P.r8['ref'] = "P"
# W.r.jm8['min.exons'] = 8
# B.r8['min.exons'] = 8
# P.r['min.exons'] = 8

All = do.call("rbind",list(W.r,B.r,P.r,W.r2,B.r2,P.r2,W.r20,B.r20,P.r20))
All %>% filter(source=="simSplit") %>% ggplot(.,aes(x=M2f,fill=as.factor(min.exons)))+geom_density(alpha=0.5)+facet_grid(cols = vars(ref))+scale_x_log10(limits=c(0.1,12))
All %>% mutate(multiIso = ifelse(parent %in% mi$V1, 1,0)) %>% group_by(ref,source,min.exons) %>% summarise(num_multIso = sum(multiIso), n = n(), prop = sum(multiIso) / n())

ALL %<>% separate(prog, sep = ",", into=c("gene1","gene2","gene3","gene4"))
  
ALL %<>% mutate(multiIso = ifelse((parent %in% mi$V1 | gene1 %in% mi$V1 | gene2 %in% mi$V1 | gene3 %in% mi$V1 | gene4 %in% mi$V1), 1, 0)) 

# MultiIsoform genes are not overrepresented in our data 
ALL %>% filter(geno=="B") %>% group_by(source) %>% summarise(num_multIso = sum(multiIso), n = n(), prop = sum(multiIso) / n())

# Multiple Isoforms does not have strong effect on null distributions
ALL$source = as.factor(ALL$source)
levels(ALL$source) <- c("Sim. Merge","Sim. Split","Candidates")
ALL %>% filter(geno == "B") %>% ggplot(.,aes(x=M2f,fill=as.factor(multiIso)))+geom_density(alpha=0.5) + scale_fill_discrete(name = "Multiple Isoforms", labels = c("Yes","No")) + scale_x_log10(limits=c(0.1,12)) + facet_grid(~source)

#No effect of genotype on null distributions
ALL %>% filter(source!="real") %>% ggplot(.,aes(x=M2f,fill=as.factor(geno)))+geom_density(alpha=0.5) + scale_x_log10(limits=c(0.1,12)) + scale_fill_discrete(name = "Genotype") + facet_grid(cols = vars(source))
ALL %>% group_by(geno, source) %>% summarize(lq = quantile(M2f, 0.1, na.rm=T), uq = quantile(M2f, 0.9, na.rm=T))

# UQuantiles (0.9) are 2.78, 2.66, and 2.97 and LQuantiles (0.1) are 1.16, 1.14, and 1.16 for B, P, and W respectively

# B is merged parent.  Not more likely to split multiIso B genes.
ALL %>% filter(source=="real" & geno !="B" & grepl("Zm00001", parent)) %>% group_by(Call) %>% summarize(n=sum(multiIso), prop = sum(multiIso)/n())
```



#Are genes that are getting split disproportionately cases where one of the split genes is a single exon?
```{r}
all = do.call("rbind", list(pp,ww,bb))
numEx = all %>% distinct(gene,numExons)
splits = ALL %>% filter(Call=="Split") %>% separate(prog,c("gene1","gene2"))
numEx['gene1']=numEx$gene
numEx['gene2']=numEx$gene

Splits=merge(Splits,numEx[,c('gene1','numExons')],by=c('gene1'))
Splits=merge(Splits,numEx[,c('gene2','numExons')],by=c('gene2'), suffixes=c(".g1",".g2"))

splits.m=melt(splits,value.var=c("numExons.g1","numExons.g2"),id.var=c("gene2","gene1","parent","source","M2f","numTiss","BiDiffMean","geno","Call"))

Splits.m %>% filter(source=="real" & Call != "NA") %>% ggplot(.,aes(x=value)) + geom_histogram() + xlab("Number of Exons in Split genes") + facet_grid(~Call, scales = "free_y")
```

Investigating quality metrics of B gene models

Position	Definition
1	Length of the 5 UTR
2	Fraction of splice sites confirmed by an EST/mRNA-seq alignment
3	Fraction of exons that match an EST/mRNA-seq alignment
4	Fraction of exons that overlap EST/mRNA-seq or protein alignments
5	Fraction of splice sites confirmed by ab initio gene prediction
6	Fraction of exons that overlap an ab initio gene prediction
7	Number of exons in the mRNA
8	Length of the 3 UTR
9	Length of the protein sequence produced by the mRNA

```{r}
an = read.table("~/Documents/Research/Maize/Split_genes/misc/B73v4.annotation_scores2.txt",head=T, fill=T)

an %<>% separate(QI,into=c("Q1","Q2","Q3","Q4","Q5","Q6","Q7","Q8","Q9"),sep = "[|]")
an %<>% separate(zscore.length.ave.sd.sp_list.ort_id.syn_id,into=c("zscore"),sep = "[:]")

#ZSCORE IS BASED ON CANONICAL TRANSCRIPT...KEEP IN MIND FOR ZSCORES INTERPRETATION
sss = ALL %>% filter(source=="real" & Call == "Split") %>% separate(prog, sep = ",", into=c("gene1","gene2","gene3","gene4"))
ddd = ALL %>% filter(source=="real" & Call == "Merged") %>% separate(prog, sep = ",", into=c("gene1","gene2","gene3","gene4"))

#can't use grepl because it will catch the simulated genes
an %<>% mutate(State = ifelse(gene_id %in% ddd$gene1 | gene_id %in% ddd$gene2 | gene_id %in% ddd$gene3 | gene_id %in% ddd$gene4 | gene_id %in% sss$gene1 | gene_id %in% sss$gene2 | gene_id %in% sss$gene3 | gene_id %in% sss$gene4, "Split", ifelse(gene_id %in% ddd$parent | gene_id %in% sss$parent, "Merged", "NoCall")), Call = ifelse(gene_id %in% sss$parent | gene_id %in% sss$gene1 | gene_id %in% sss$gene2 | gene_id %in% sss$gene3 | gene_id %in% sss$gene4, "Split", ifelse(gene_id %in% ddd$parent | gene_id %in% ddd$gene1 | gene_id %in% ddd$gene2 | gene_id %in% ddd$gene3 | gene_id %in% ddd$gene4, "Merged", "NoCall")))

an %<>% mutate(ISPAR = ifelse(gene_id %in% bb[bb$source=="real",]$parent, TRUE, ifelse(gene_id %in% pp[pp$source=="real",]$parent, TRUE, ifelse(gene_id %in% ww[ww$source=="real",]$parent, TRUE, FALSE))))

an %<>% mutate(State = ifelse(Call=="NoCall" & gene_id %in% OO5$gene, "One-to-one", State))

an %<>% mutate(group = paste(Call,State,sep=""))

an$group = as.factor(an$group)
an$group = factor(an$group,levels(an$group)[c(4,6,5,1,2,3)])

an %>% filter(zscore!="ND") %>% ggplot(.,aes(fill=Call,y=as.numeric(zscore))) + geom_boxplot() + ylab("Z-score of diff. in length b/n orthologous proteins") + theme_bw() + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

#FILTER FOR MAX AED ACROSS ALL TRANSCRIPTS
an %>% filter(State != "NoCall") %>% mutate(Genotype = "B") %>% group_by(group, gene_id, Genotype) %>% summarize(AED = max(AED)) %>% ggplot(., aes(x = Genotype, fill = group, y = as.numeric(AED))) + geom_boxplot(outlier.shape=NA) + ylab("Annotation Edit Distance (AED)")  + xlab("Genotype") + theme_bw() + scale_fill_manual(name = "Annot./Called", labels = c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge", "Merge/Split"), values = wes_palette("Royal2")) + theme(axis.text.x = element_text(size = 16), axis.title.x = element_text(size = 18), axis.text.y = element_text(size = 16), axis.title.y = element_text(size = 18), legend.text = element_text(size = 16), legend.title = element_text(size = 18))

#Maker Quality 
an %>% ggplot(.,aes(fill=Call,y=as.numeric(Q9))) + geom_boxplot() + ylab("Protein sequence length according to mRNA") + theme_bw() + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
an %>% ggplot(.,aes(fill=Call,y=as.numeric(Q1))) + geom_boxplot() + scale_y_log10() + ylab("Length of the 5 UTR") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
an %>% ggplot(.,aes(fill=Call,y=as.numeric(Q2))) + geom_boxplot() + scale_y_log10() + ylab("Prop. splice sites confirmed by EST/mRNA-seq alignment") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
an %>% ggplot(.,aes(fill=Call,y=as.numeric(Q3))) + geom_boxplot() + scale_y_log10() + ylab("Prop. exons matching EST/mRNA-seq alignment") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
an %>% ggplot(.,aes(fill=Call,y=as.numeric(Q4))) + geom_boxplot() + scale_y_log10() + ylab("Prop. exons overlapping EST/mRNA-seq/protein alignments") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
an %>% ggplot(.,aes(fill=Call,y=as.numeric(Q5))) + geom_boxplot() + scale_y_log10() + ylab("Prop. splice sites confirmed by ab initio prediction") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
an %>% ggplot(.,aes(fill=Call,y=as.numeric(Q6))) + geom_boxplot() + scale_y_log10() + ylab("Prop. exons overlapping ab initio prediction") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
an %>% ggplot(.,aes(fill=Call,y=as.numeric(Q7))) + geom_boxplot() + scale_y_log10() + ylab("Number of exons in the mRNA") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
an %>% ggplot(.,aes(fill=Call,y=as.numeric(Q8))) + geom_boxplot() + scale_y_log10() + ylab("Length of the 3 UTR") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())


#PH207 AED scores
aed = read.table("~/Documents/Research/Maize/Split_genes/misc/PH207_AEDs.txt")
nom = read.table("~/Documents/Research/Maize/Split_genes/misc/PH207_nomenclature_transformation.txt")
nom['key'] = nom$V3
aed['key'] = aed$V1
AED = merge(aed, nom, by = "key")
AED['gene_id'] = AED$V5

AED %<>% mutate(ISPAR = ifelse(gene_id %in% bb[bb$source=="real",]$parent, TRUE, ifelse(gene_id %in% pp[pp$source=="real",]$parent, TRUE, ifelse(gene_id %in% ww[ww$source=="real",]$parent, TRUE, FALSE))))

AED %<>% mutate(State = ifelse(gene_id %in% ddd$gene1 | gene_id %in% ddd$gene2 | gene_id %in% ddd$gene3 | gene_id %in% ddd$gene4 | gene_id %in% sss$gene1 | gene_id %in% sss$gene2 | gene_id %in% sss$gene3 | gene_id %in% sss$gene4, "Split", ifelse(gene_id %in% ddd$parent | gene_id %in% sss$parent, "Merged", "NoCall")), Call = ifelse(gene_id %in% sss$parent | gene_id %in% sss$gene1 | gene_id %in% sss$gene2 | gene_id %in% sss$gene3 | gene_id %in% sss$gene4, "Split", ifelse(gene_id %in% ddd$parent | gene_id %in% ddd$gene1 | gene_id %in% ddd$gene2 | gene_id %in% ddd$gene3 | gene_id %in% ddd$gene4, "Merged", "NoCall")))

AED %<>% mutate(State = ifelse(Call=="NoCall" & gene_id %in% OO5$gene, "One-to-one", State))

AED %<>% mutate(group = paste(Call,State,sep=""))

AED$group = as.factor(AED$group)
AED$group = factor(AED$group,levels(AED$group)[c(4,6,5,1,2,3)])

AED %>% filter(State != "NoCall") %>% mutate(Genotype = "B") %>% group_by(group, gene_id, Genotype) %>% summarize(AED = max(V2.x)) %>% ggplot(., aes(x = Genotype, fill = group, y = as.numeric(AED))) + geom_boxplot(outlier.shape=NA) + ylab("Annotation Edit Distance (AED)")  + xlab("Genotype") + theme_bw() + scale_fill_manual(name = "Annot./Called", labels = c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge", "Merge/Split"), values = wes_palette("Royal2")) + theme(axis.text.x = element_text(size = 16), axis.title.x = element_text(size = 18), axis.text.y = element_text(size = 16), axis.title.y = element_text(size = 18), legend.text = element_text(size = 16), legend.title = element_text(size = 18))

AED1 = AED %>% filter(State != "NoCall") %>% mutate(Genotype = "P") %>% group_by(group, gene_id, Genotype) %>% summarize(AED = max(V2.x))
AN = an %>% filter(State != "NoCall") %>% mutate(Genotype = "B") %>% group_by(group, gene_id, Genotype) %>% summarize(AED = max(AED))

AN = rbind(AED1,AN)

AN %<>% group_by(Genotype, group) %>% mutate(outlier = (AED > quantile(AED, 0.75) + IQR(AED) * 1.5)) %>% ungroup()

#Violin plus box
AN %>% ggplot(.,aes(x=Genotype,fill=group,y=AED)) + geom_violin() + geom_boxplot(outlier.shape=NA, width=0.2, position=position_dodge(width=0.9)) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_color_manual(guide=F,name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + xlab("Genotype") + ylab("Annotation Edit Distance (AED)") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

#Box plus point outliers
AN %>% ggplot(.,aes(x=Genotype,fill=group,y=AED)) + geom_boxplot(outlier.shape=NA) + geom_point(data = AN, aes(x=Genotype,fill=group,y=AED, alpha=outlier,color=group), position=position_jitterdodge(dodge.width=0.8)) + scale_alpha_discrete(guide = F, range=c(0,0.4)) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_color_manual(guide=F,name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + xlab("Genotype") + ylab("Annotation Edit Distance (AED)") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

#Just Box
AN %>% ggplot(.,aes(x=Genotype,fill=group,y=AED)) + geom_boxplot(outlier.shape=NA) + scale_fill_manual(name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_color_manual(guide=F,name="Annot./Called",labels=c("One-to-one", "Split/Split", "Split/Merge", "Merge/Merge","Merge/Split"),values=wes_palette("Royal2")) + scale_x_discrete(breaks = c("B", "P"), labels = c("B73", "PH207")) + xlab("Annotation") + ylab("Annotation Edit Distance (AED)") + theme_bw() + theme(axis.text.y=element_text(size=16), axis.title.y=element_text(size=18),axis.text.x=element_text(size=16),axis.title.x=element_text(size=18),legend.text = element_text(size=16), legend.title = element_text(size=18))

```

Correlation in M2f and consistency in calls in overlapping candidate sets
E.g. 1W gene = 2B genes and same 1W gene = 2P genes
```{r}
ALL = do.call("rbind", list(P.r, B.r, W.r))
ALL %<>% mutate(Call = ifelse(M2f > quantile(ALL[ALL$source=="simSplit",]$M2f,0.9,na.rm=T), "Split", ifelse(M2f < quantile(ALL[ALL$source=="simMerged",]$M2f,0.1,na.rm=T), "Merged", "NoCall")))

ALL.r=ALL%>% filter(source=="real")

dub.pars = ALL.r %>% group_by(parent) %>% summarize(n=n()) %>% filter(n==2)
Dub.pars = ALL.r %>% filter(parent %in% dub.pars$parent) %>% mutate(dubtype = "merged")
Dub.pars %<>% mutate(Geno = ifelse(grepl("Zm00001",prog),"B",ifelse(grepl("Zm00008",prog),"P","W")), key = parent)
dub.prog = ALL.r %>% group_by(prog) %>% summarize(n=n()) %>% filter(n==2)
Dub.prog = ALL.r %>% filter(prog %in% dub.prog$prog) %>% mutate(dubtype = "splits")
Dub.prog %<>% mutate(Geno = ifelse(grepl("Zm00001",parent),"B",ifelse(grepl("Zm00008",parent),"P","W")), key = prog)

dubs = rbind(Dub.pars,Dub.prog)
dubs %<>% mutate(Par = ifelse(grepl("Zm00001",parent),"B",ifelse(grepl("Zm00008",parent),"P","W")), Call = ifelse(M2f > quantile(ALL[ALL$source=="simSplit",]$M2f,0.9,na.rm=T), "Split", ifelse(M2f < quantile(ALL[ALL$source=="simMerged",]$M2f,0.1,na.rm=T), "Merged", "NoCall")))

dubs.m = melt(dubs, id.vars=c("parent","prog","source","numTiss","BiDiffMean","dubtype","Geno","key","Par"))

Dubs = dubs.m %>% dcast(.,key + dubtype ~ Geno + variable,value.var=c("value"))
Dubs %>% filter(dubtype=="merged") %>% ggplot(.,aes(x=W_M2f, y = P_M2f)) + geom_point() + geom_smooth(method="lm")

#change to numeric
Dubs$P_M2f = as.numeric(Dubs$P_M2f)
Dubs$W_M2f = as.numeric(Dubs$W_M2f)
Dubs$B_M2f = as.numeric(Dubs$B_M2f)

#plot
Dubs %>% filter(dubtype=="merged") %.>% ggplot(.) + geom_point(data=.,aes(y = P_M2f, x = B_M2f, color="#E1BD6D"), alpha=0.5)+ geom_smooth(data=.,aes(y = P_M2f, x = B_M2f, color = "#E1BD6D"), method = "lm", se = F) + geom_point(data=.,aes(y = B_M2f, x = W_M2f, color = "#0B775E"), alpha=0.5)+ geom_smooth(data=.,aes(y = B_M2f, x = W_M2f, color = "#0B775E"), method = "lm", se = F) + geom_point(data=.,aes(y = P_M2f, x = W_M2f, color = "#35274A"), alpha=0.5) + geom_smooth(data=.,aes(y = P_M2f, x = W_M2f, color = "#35274A"), method = "lm", se = F) + geom_hline(yintercept = 2.809204, color = "#619CFF", linetype = "dashed")  + geom_vline(xintercept = 2.809204, color = "#619CFF", linetype = "dashed") + geom_hline(yintercept = 1.156629, color = "#F8766D", linetype = "dashed")  + geom_vline(xintercept = 1.156629, color = "#F8766D", linetype = "dashed") + scale_color_manual(name = "Comparison", values = c("#E1BD6D", "#0B775E", "#35274A"), labels=c("B/W", "P/W", "P/B")) + theme_bw() + xlab(expression(paste("M2f"[1]))) + ylab(expression(paste("M2f"[2]))) + ylim(0,10) + xlim(0,10)

Dubs %>% filter(dubtype=="merged") %.>% ggplot(.) + geom_hline(yintercept = 2.809204, color = "#619CFF", linetype = "dashed")  + geom_vline(xintercept = 2.809204, color = "#619CFF", linetype = "dashed") + geom_hline(yintercept = 1.156629, color = "#F8766D", linetype = "dashed")  + geom_vline(xintercept = 1.156629, color = "#F8766D", linetype = "dashed") + theme_bw() + xlab(expression(paste("M2f"[1]))) + ylab(expression(paste("M2f"[2]))) + ylim(0,10) + xlim(0,10)

```

Correlation in expression.
Investigating whether expression estimates in one-to-one genes are more highly correlated as compared to the split-gene candidates, particularly those that were classified as split or merged.
```{r}

# One-to-one genes
oo = read.table("~/Documents/Research/Maize/Split_genes/data/One_to_ones.txt")
oo['gene'] = oo$V1
oo %<>% group_by(V1,V2) %>% mutate(gene1 = min(as.character(V1),as.character(V2)), gene2 = max(as.character(V1),as.character(V2))) # Sort the one-to-one gene ids so that we can find reciprocals as pairs of matching gene ids

#Filter for only reciprical one-to-one
OO = oo %>% group_by(gene1,gene2) %>% summarize(n=n()) %>% filter(n==2)

#What about filter for genes that are reciprical one-one across all 3 comparisons?
OO3 = oo %>% group_by(gene1) %>% summarize(n=n()) %>% filter(n==4) %>% mutate(gene = gene1) %>% select(gene)
OO4 = oo %>% group_by(gene2) %>% summarize(n=n()) %>% filter(n==4) %>% mutate(gene = gene2) %>% select(gene)
OO5 = rbind(OO3, OO4)

OO6 = oo %>% filter(V1 %in% OO5$gene | V2 %in% OO5$gene) %>% select(V1,V2) %>% gather(value = "gene") %>% distinct(gene)

OO2 = oo %>% filter(V1 %in% OO$gene1 | V1 %in% OO$gene2) %>% select(V1, V2, gene)

#Calculate normalized expression and format data
pp2 = formatData(p.counts, p.splits, p.fsplits, fmt = "m2f")
ww2 = formatData(w.counts, w.splits, w.fsplits, fmt = "m2f")
bb2 = formatData(b.counts, b.splits, b.fsplits, fmt = "m2f")
aa = do.call("rbind", list(pp2, ww2, bb2))

#Average across replicates.  These will be x-values for scatterplot
AA = aa %>% group_by(gene, tissue, sample) %>% summarize(value = mean(value))
AA['gene1'] = AA$gene
AA1 = merge(AA, OO2, by="gene")
AA1['gene2'] = AA1$V2

#Trying to match up x-values with y-values here, which would be expression of the partner in the one-to-one relationship 
AA2 = AA1
AA2['gene3'] = AA2$gene1 # Store gene1 value
AA2['gene1'] = AA2$gene2 #Flip gene1 and gene2
AA2['gene2'] = AA2$gene3

AA1 %<>% select(tissue,sample,value,gene2, gene1)
AA2 %<>% select(tissue,sample,value,gene2, gene1) # Get rid of extra variables

AA3 = merge(AA1, AA2, by=c("gene1","gene2","tissue", "sample"))

#Why are there so many one-to-one genes with no expression in one.
AA3 %>% sample_n(100000) %>% ggplot(.,aes(y=value.y, x = value.x)) + geom_point() + scale_x_log10() + scale_y_log10()

#Correlation in expression for split-gene candidates
aa.prog = aa %>% filter(source == "real" & isParent==FALSE) %>% group_by(gene, tissue, sample) %>% summarize(value = mean(value))

jj = aa %>% group_by(gene, tissue, sample) %>% summarize(value = mean(value))
jj['parent'] = jj$gene

kk = merge(aa.prog,jj, by=c("gene","tissue","sample"))

# Currently y is merged genes and x axis is for split genes.  Results in unbalanced scatterplot.  Here im just reflecting the points so that they are represented twice in dataset
ll = kk
ll['a'] = kk$value.x
ll['b'] = kk$value.y
ll$value.y = ll$a
ll$value.x = ll$b
ll %<>% select(-a, -b)

KK = rbind(ll, kk)

#downsample one-to-ones so that data is of equivalent size to split-gene candidates.  
AA4 = AA3 %>% sample_n(nrow(kk))

#Plot correlation
ggplot() + geom_point(data = KK, aes(y=value.y, x = value.x, color = "blue"), alpha=0.2) + geom_point(data=AA4, aes(x=value.x, y = value.y, color = "green"), alpha=0.2) + scale_color_manual(name = "", values = c("blue","green"), labels = c("Split-gene\nCandidates", "1-to-1")) + scale_x_log10(name = "Gene 1 Normalized Expression (TPM)", label = comma) + scale_y_log10(name = "Gene 2 Normalized Expression (TPM)", label = comma) + theme_bw()

kk %>% filter(value.y!=0 & value.x!=0) %.>% cor(.$value.x,.$value.y)
AA4 %>% filter(value.y!=0 & value.x!=0) %.>% cor(.$value.x,.$value.y)



```

# 3' analysis (TEnders data)
```{r}
## THIS IS UNFILTERED DATA
p3 = read.table("~/Documents/Research/Maize/Split_genes/data/TEnders_3prime_cln.txt", head =T)
b.SM = ALL %>% filter(grepl("Zm00001", prog) & source=="real" & Call == "Merged")
b.SM %<>% separate(prog, sep = ",", into=c("gene1","gene2","gene3","gene4"))
p3.b = p3 %>% filter(Geneid %in% b.SM$gene1 | Geneid %in% b.SM$gene2 | Geneid %in% b.SM$gene3 | Geneid %in% b.SM$gene4)
p3.b.m = melt(p3.b, id.var = "Geneid")

p3.b.m %<>% separate(variable, sep = "[.]", into = c("sample","bam","treatment")) %>% select(Geneid,Length,sample, treatment, value) %>% separate(sample, sep = "_", into = c("d","sample")) %>% select(Geneid,Length,sample, treatment, value)

p3.b.m %<>% mutate(gene1 = Geneid, gene2 = Geneid, gene3 = Geneid, gene4 = Geneid)

p1 = p3.b.m %>% inner_join(., b.SM, by = "gene1") %>% select(Geneid, Length, sample, treatment, value, parent, numTiss, M2f, Call, Par)
p2 = p3.b.m %>% inner_join(., b.SM, by = "gene2") %>% select(Geneid, Length, sample, treatment, value, parent, numTiss, M2f, Call, Par) 
p3 = p3.b.m %>% inner_join(., b.SM, by = "gene3") %>% select(Geneid, Length, sample, treatment, value, parent, numTiss, M2f, Call, Par)
p4 = p3.b.m %>% inner_join(., b.SM, by = "gene4") %>% select(Geneid, Length, sample, treatment, value, parent, numTiss, M2f, Call, Par)

p3.j = do.call("rbind", list(p1, p2, p3, p4))
p3.j.s = p3.j %>% group_by(parent, Geneid, M2f, Call, numTiss) %>% summarize(exp = mean(value, na.rm=T))

p3.j.s %>% group_by(parent) %>% summarize(min = min(exp), max = max(exp)) %>% melt(., id.var = "parent") %>% ggplot(., aes(x = variable, y = value)) + geom_boxplot() + scale_y_log10()

bams = list.files("~/Documents/Research/Maize/Split_genes/data/3prime/", pattern ="*bam$", full.names = T)
fc = featureCounts(files = bams, annot.ext = "~/Documents/Research/Maize/references/Zea_mays.AGPv4.33.genes_only.gff3", isGTFAnnotationFile = T, GTF.featureType="gene")
fcc = as.data.frame(fc$counts)
fcc['gene'] = rownames(fcc)
fcc = melt(fcc, id.var = "gene")
fcc %<>% separate(variable, "[.]", into = c("a", "b","c","d","e","f","g","h","i","sample","k")) %>% select(gene, sample, value)
fcc1 = fcc %>% filter(gene %in% b.SM$gene1 | gene %in% b.SM$gene2 | gene %in% b.SM$gene3 | gene %in% b.SM$gene4)

fcc1 %<>% mutate(gene1 = gene, gene2 = gene, gene3 = gene, gene4 = gene)

p1 = fcc1 %>% inner_join(., b.SM, by = "gene1") %>% select(gene, sample, value, parent, numTiss, M2f, Call)
p2 = fcc1 %>% inner_join(., b.SM, by = "gene2") %>% select(gene, sample, value, parent, numTiss, M2f, Call)
p3 = fcc1 %>% inner_join(., b.SM, by = "gene3") %>% select(gene, sample, value, parent, numTiss, M2f, Call)
p4 = fcc1 %>% inner_join(., b.SM, by = "gene4") %>% select(gene, sample, value, parent, numTiss, M2f, Call)

fcc2 = do.call("rbind", list(p1, p2, p3, p4))
fcc2.s = fcc2 %>% filter(grepl("B73", sample)) %>% group_by(parent, gene, M2f, Call, numTiss) %>% summarize(exp = mean(value, na.rm=T))

fcc2.s %>% group_by(parent) %>% summarize(min = min(exp), max = max(exp)) %>% melt(., id.var = "parent") %>% ggplot(., aes(x = variable, y = value)) + geom_boxplot() + scale_y_log10()
fcc2.s %>% group_by(parent) %>% summarize(min = min(exp), max = max(exp)) %>% ggplot(., aes(x = min, y = max)) + geom_point() + scale_x_log10() +scale_y_log10()

```

